---
title: "lab2"
format:
  html:
    self-contained: true
---

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(forcats)
```

#Question1

```{r}
trees <- read_csv("data/tree_dat.csv", show_col_types = FALSE)
glimpse(trees)
```

#The tibble has r nrow(trees) rows and r ncol(trees) columns:treeID (int), standID (int), stand (chr), year (int; 1897–2007),species (chr; 15 codes), age (dbl), inc (dbl mm), rad_ib (dbl mm).

#Question 2 #There are r stand1_count records from stand 1.

```{r}
stand1_count <- trees %>%
  filter(standID == 1) %>%
  nrow()
stand1_count
```

#Question 3

```{r}
abba_pist_count <- trees %>%
  filter(species %in% c("ABBA","PIST")) %>%
  count(species)
abba_pist_count
```

#2 records

#Question 4

```{r}
last_year <- max(trees$year)
over200_last <- trees %>%
  filter(year == last_year, age > 200) %>%
  nrow()
over200_last
```
# 7 trees

#Question 5
```{r}
oldest_tree <- trees %>%
  slice_max(age, n = 1)
oldest_tree
```
# 269 years old

#Question 6
```{r}
old5_2001 <- trees %>%
  filter(year == 2001) %>%
  slice_max(age, n = 5)
old5_2001
```
#Question 7
```{r}
n_2002      <- trees %>% filter(year==2002) %>% nrow()
sample30_2002 <- trees %>%
  filter(year == 2002) %>%
  slice_sample(prop = 0.30) %>%
  nrow()
sample30_2002
```
# 687

#Question 8
```{r}
top3_stand5_2007 <- trees %>%
  filter(standID == 5, year == 2007) %>%
  arrange(desc(rad_ib)) %>%
  slice_head(n = 3) %>%
  pull(treeID)
top3_stand5_2007
```
# 128, 157, 135

#Question 9
```{r}
small3_stand3_2007 <- trees %>%
  filter(standID == 3, year == 2007) %>%
  select(treeID, stand, year, rad_ib) %>%
  slice_min(rad_ib, n = 3)

small3_stand3_2007
```
#Question 10
```{r}
trees_no_stand <- trees %>%
  select(-stand)
glimpse(trees_no_stand)
```
#Question 11
```{r}
id_cols <- trees %>%
  select(contains("ID"))
glimpse(id_cols)
```
#names is one of them

#Question 12
```{r}
id_or_stand <- trees %>%
  select(matches("ID|stand"))
glimpse(id_or_stand)
```

#Question 13
```{r}
trees_renamed <- trees %>%
  rename(
    rad_ib_mm = rad_ib,
    inc_mm    = inc
  )
glimpse(trees_renamed)
```

#Question 14
```{r}
trees_ba <- trees_renamed %>%
  mutate(
    DBH_cm = (rad_ib_mm * 2) / 10,
    BA_m2  = 0.00007854 * DBH_cm^2
  )

potr_2007_mean_ba <- trees_ba %>%
  filter(species == "POTR", year == 2007) %>%
  summarise(mean_BA = mean(BA_m2, na.rm = TRUE))
potr_2007_mean_ba
```
# 0.03696619	

#Question 15
```{r}
trees_est <- trees %>%
  mutate(established = age > 5) %>%
  count(established)
trees_est
```

# FALSE	8883			
# TRUE	122503	

#Question 16
```{r}
trees_class <- trees_ba %>%
  mutate(
    DBH_class = case_when(
      DBH_cm <=  2.5 ~ "seedling",
      DBH_cm <= 10   ~ "sapling",
      DBH_cm <= 30   ~ "pole",
      TRUE           ~ "sawlog"
    )
  ) %>%
  filter(year == 2007) %>%
  count(DBH_class)
trees_class
```
# pole	1963			
# sapling	252			
# sawlog	76	

#Question 17
```{r}
dbh_stats_2007 <- trees_ba %>%
  filter(year == 2007) %>%
  summarise(
    mean_DBH = mean(DBH_cm, na.rm = TRUE),
    sd_DBH   = sd(DBH_cm,   na.rm = TRUE)
  )
dbh_stats_2007
```
# The mean is 16.09351cm and the standard deviation is 6.138643cm

#Question 18
```{r}
oldest_species_2003 <- trees %>%
  filter(year == 2003) %>%
  group_by(species) %>%
  summarise(mean_age = mean(age, na.rm = TRUE)) %>%
  arrange(desc(mean_age)) %>%
  slice_head(n = 3)
oldest_species_2003
```
# species
#<chr>
#mean_age
#<dbl>
#THOC	126.63830			
#FRNI	83.08333			
#PIST	73.28571	

# Question 19
```{r}
year_summary <- trees %>%
  summarise(
    unique_years   = n_distinct(year),
    first_year     = min(year),
    last_year      = max(year)
  )
year_summary
```
# unique_years
#<int>
#first_year
#<dbl>
#last_year
#<dbl>
#111	1897	2007	

#Question 20

```{r}
stand_years <- trees %>%
  group_by(stand) %>%
  summarise(n_years = n_distinct(year)) 

max_years <- max(stand_years$n_years)

stands_longest <- stand_years %>%
  filter(n_years == max_years)
stands_longest
```
# Final Question 

```{r}
growth_rates <- trees %>%
  arrange(treeID, year) %>%
  group_by(treeID, species) %>%
  mutate(
    yr_growth = inc - lag(inc)
  ) %>%
  filter(!is.na(yr_growth))

species_stats <- growth_rates %>%
  group_by(species) %>%
  summarise(
    mean_growth = mean(yr_growth, na.rm = TRUE),
    sd_growth   = sd(yr_growth,   na.rm = TRUE)
  )

fastest <- species_stats %>% arrange(desc(mean_growth)) %>% slice_head(n = 3)
slowest <- species_stats %>% arrange(mean_growth)          %>% slice_head(n = 3)

fastest
slowest
```


